\chapter*{Acknowledgements}

First of all, I would like to thank Liva and Stéphane for reviewing my thesis (in a short time).
Moreover, I would like to thank Marc and Rémi for accepting to be, respectively, the examiner and the president; I am very honored to have you on my jury!
\\

I would like to sincerely thank my supervisors, who were always here and patient with me; I believe that I was very lucky!
Thank you, Amaury, for proposing this PhD subject to me and for always having a critical viewpoint on my work.
In the meantime, I always felt free to work where and when I want; thank you.
Thank you, Pascal, for your guidance when writing the papers. 
Our discussions were always interesting, and I learned a lot.
Visiting Laval University has not been possible with the pandemic\dots I hope that it will be possible later!
Thank you, Emilie, for being there literally every day for me. 
I really enjoyed talking about science and personal things with you.
Thank you for showing me (for a long time) how to write a scientific paper and how to do presentations.
Moreover, thank you for letting me the chance to perform lectures (which was hard but so interesting).
Lastly, thanks, Emilie and Cécile, for making me discover Kung Fu (which I gave up but enjoyed a lot)!
Thanks, Valentina and Rémi, for having contributed together. 
This thesis would not have looked like this without you.\\

This PhD journey has also been full of fun moments and scientific discussions among the permanent and non-permanent staff of the lab.
Generally, thank you all\footnote{The list is too long to be enumerated!} for the good mood in the lab that I really enjoyed during these years.
More specifically, thanks, Léo and Jules, for introducing me to the ``ASEC'' parties that were awesome with you!
I also want to thank the ``ASEC'' team (\ie, Aubin, Damien, Jules, and, Léo) for the good nights.
Unfortunately, the pandemic\footnote{Thanks, Brrrmmaou, Titi, and Bibi, for sleeping next to me when I was working; it was motivating!} stopped, for a long time, these good moments\dots\\

Moreover, this thesis marks the end of all those school years. 
I would have never bet on myself about writing a ``book'', especially in English!
This could never have been done if I had stopped studying early; thanks, mom and dad, for your support during these years.
More generally, I want to thank my family, who have all been supportive during these moments of hard work with the projects, the exams, and during the PhD.\\

I want to thank some special friends that have been important to me since primary, middle, high school, and the university.
Thanks, Jules, for all these years: I would never have been the same without you! 
Thank you for making me discover the Loire gorges along with hiking since middle school. 
Our philosophical discussions in this area are so memorable! 
Your recklessness helped me a lot to push my limits, \eg, during the urban and cave explorations.
Thank you for accompanying me in my first climbing and paragliding!
Thanks, Fleurian, for these school years: our hiking and climbing sessions were so cool (and usually with Jules)! 
Moreover, thank you for making me discover a new sport while writing my thesis: karting.
I hope that there will be other road trips with the Dyane car!
Thanks, Benoit, for sharing my passion for computer science and coding together for all these years. 
For instance, we coded an optimization algorithm in high school!
Likewise, thanks, Angélique, for sharing your interest in computer science with me since high school.
I hope that we will continue to hike or drink tea at the Oriol Tower!
Thanks, Guillaume, for these wonderful student years spent with you since the first year of ``licence''!
We shared many parties, met our respective friends, and had good times.
Moreover, thank you for doing sports with me, such as climbing, ice skating, and hiking (during classes).
These good moments were accompanied by a lot of work: for you for being my project partner for almost five years! 
I also appreciated our first internship together at the same desk and our NeurIPS paper! 
It is funny where life leads us.
I would like to thank Coralie for asking me for information about horse riding during my first internship.
Since then, I have ridden again and learned a lot about horses!
Your restlessness (and your desire to always do something) has been valuable, especially when avoiding thinking about the thesis.\\

Last but not least, thank you, Chloé, for your unwavering support over these past two years.
It was precious, notably for the difficult moment like the submissions and the thesis writing!
Thank you for removing (sometimes) my nose from the grindstone to enjoy life (with you).
I really hope to continue horse riding, hiking, traveling, or having common projects. 